Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-08-22 Multiresolution_framework/ltx/response_to_reviews.tex

\emph{We understand that this section will not be the central focus of the paper for some members of our research community, however we believe that it is important that this information remains in the paper to allow for implementation without the requirement of digesting information from several texts.}

\emph{We have had similar feedback to your suggestion from our colleagues prior to submission of the paper. Interestingly, the feedback from the people with expertise in wavelet theory thought that the overview of the estimation methods was essential, but the section on wavelets should be reduced. Conversely, people with expertise in the estimation methods thought the overview on wavelets was critical and the section on EM-algorithm should be reduced. We completely appreciate R1's comments and suggestions, but since R2 commented on the usefulness of this section we would very much like to leave it in. We believe that different sections will be valuable for people with different specializations, the synthesis of this information is a valuable contribution for implementation for a wide audience.}

        \item Section 6 also borrows heavily from Dewar et al, e.g. cf equation (58) of this paper with equation (19) of Dewar et al: the state transition matrix is a perturbation of the corresponding matrix used in that earlier paper. The approach used in both papers seems to be the same with an RTS smoother used in the E-step (this was also used and explained in Freestone et al.) and an M-step optimizing a cost function, quadratic in the parameters, cf. equation (62) with equation (29) of Theorem 2 of Dewar et al. In general, the derivations of section 6 and appendix B follow closely or are very similar to those of Dewar et al.

		\emph{To clarify the advantage of the M-step over the implementation in Dewar et al. we have added a paragraph to Section 8.1 (now 8.2 ``Estimation Algorithm''). Further comments for the reviewer are added below.}

@@

\emph{The inverse framework in Dewar et al. (2009) is based on a general integro-differential equation model, which is not suitable when neural field models are considered. As a result, the necessary modifications have been made. }

\emph{This paper has a novel implementation of the M-step, which significantly simplifies the computational complexity of the EM-based algorithm proposed by Dewar et al. (2009). These modifications are significant and crucial for high-dimensional systems. More specifically, implementing the multi-resolution framework introduced in the paper would be impossible without this new contribution. The differences are highlighted in Table.~4 of the manuscript, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$, see equation (64) of this paper and equation (31) of Dewar et al. (2009), where the static and dynamic components are separated by defining $\boldsymbol \Gamma_2$ and $\mathbf U$ (see equations (66) and (67)). A simplification is also performed  (by defining $\boldsymbol \Gamma_1$, i.e. equation (65)) to calculate $\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$ (see equations (62) and (63) of this paper and equation (30) of Dewar et al. (2009)).}

@@

				% \emph{\parham{some notes to be added to the manuscript: Parsimony: as few as necessary variables to describe a phenomenon, complexity: to add detail description to the model to increase its accuracy with the natural phenomenon, finding a balance between complexity and parsimony is crucial. By parsimony we sacrifice realism ...  }}

		\emph{To clarify this comment, a short paragraph has been included in Section 6 to better explain the modeling choices facing every user. That is, choosing between a more parsimonious or more complex model structure.  For the audience with less experience in system identification, a reference with further information regarding this topic has also been included.}

		\item The comments on large scale networks and delays made on p.20 look irrelevant to the present framework.

		\emph{We have omitted the last paragraph of the discussion on delays in large scale networks as the reviewer suggested.}

@@

    \section{Reviewer 2}

\begin{enumerate}
    \item As a minor comment, I would suggest changing the notation for the scaling and wavelet coefficients in Eq. (6), for example into normal lowercase `a' and `c', respectively. The $\beta$ might be confused by B-spline notation.

	\emph{We have changed the scaling and wavelet coefficients into normal lowercase `a' and `c', respectively.}

\item For the discussion, it might be useful to add a sentence that the basis structure (i.e., non-redundant, versus a (tight) frame representation) is a requirement for this framework. A small comment might also be considered for Sect 8.2: it is possible to apply other than dyadic subsampling schemes in multiple dimensions, which leads to less wavelet subbands. For example, in 2D, the quincunx scheme is well-known to have better isotropic properties. Unfortunately, in 3D, it is not possible to have a subsampling operator (acting on a Cartesian sampling lattice) that is performing subsampling at equal rate in each dimension at each iteration, see Van De Ville et al,`On the multidimensional extension of the quincunx subsampling lattice', IEEE SPL, 12:2, pp.  112-115, 2005, for a comment on this.

\emph{We have added a paragraph  to Section~8.2 (now 8.3 ``Modification to Higher Dimensions'') regarding this important point.}

### 2012-08-22 Multiresolution_framework/ltx/response_to_reviews.tex

\end{enumerate}

    \section{Reviewer 1}

    \begin{enumerate}
        \item The exposition of the material and the relevant discussion have many similarities with the papers of the same authors (Freestone et al., 2011) and (Dewar et al., 2009) published in NeuroImage and IEEE Trans on Signal Processing  respectively. The authors say the paper at hand is an ``extension'' of Freestone et al. by which I understand they mean the use of a wavelet as opposed to a Gaussian basis for the decomposition. Since the methodology used is the same as in that earlier NeuroImage paper, the paper at hand repeats several of  the results of that previous paper, see sections 3 and 4: equations (2)-(5) and (20) -(33) have already appeared or are very similar to equations in Freestone et al.

	\emph{Freestone et al. (2011), Dewar et al. (2009) both propose solutions to representing a spatiotemporal field with a finite number of basis functions. Together with this paper, we are building tools that deal with increasing levels of complexity in the spatiotemporal dynamics. The first paper describes key work in the PhD thesis of Dewar, which introduces what we now call a dynamic Galerkin projection. The second in the PhD thesis of Freestone, which recast the estimation problem in the context of neural fields. This paper covers some of the key results from the PhD thesis of Aram, which details methods for achieving a more accurate representation of the neural fields and an elegant estimation scheme. Each of these steps was a major undertaking and should not be underestimated. There are similarities with the frameworks with all of these papers, where the field is represented by a basis function decomposition. The challenges in forming a multi-resolution state-space model included the conceptualization of the problem, the formulation and derivation of the state-space model, and innovations that make dealing with such a large-scale system feasible. This paper is a result of a concerted effort from our team over a time period of more than two years work. The progress we have made with the earlier papers has made this paper possible, just as we hope that this paper will inspire future work in this direction.}

@@

\emph{The authors would like to point out that the derivation of the state-space multi-resolution model was made general for all possible activation functions. This derivation to represent different spatial scales in the neural field simultaneously in a state-space format, such that estimation algorithms can be applied, is perhaps the most important contribution of the paper. We understand that all physical systems are nonlinear. However, we also understand that linear approximations are often required in engineering for useful models. In the introduction to the estimation section, we have attempted to discuss this issue with regard to forward vs. inverse modeling. We believe that the linear model is more acceptable for inverse modeling, but understand a nonlinear activation function is required for forward modeling. For example, in a forward model, the nonlinearity is required to prevent negative firing rates. However, in the inverse case, the physiology naturally restricts firing rates to be positive. Furthermore, in the mean field case, we do not expect firing rates of all neurons to be zero or all neurons to be simultaneously firing at their maximal rate. Therefore, a linear model may suffice to image a multi-resolution neural field and physiological parameters using electrophysiological data.}

\emph{Given restrictions of computational complexity, the choice of whether to use the framework presented in Freestone et al. or the current paper boils down to a trade-off between temporal accuracy and spatial accuracy. Furthermore, the proven convergence of estimation algorithms for linear systems may actually lead to more accurate models given a closed-loop scheme. This is due to the guaranteed convergence of the EM-algorithm for (observable) linear systems. With convergence, modeling errors can be corrected using data.}

        \item The authors say that a multi-resolution approach would explain the``organizational level of the brain at which cognition and behavior are best explained'' but it is unclear what the alternative levels considered here are and what the significance of the results obtained is.

@@

    \section{Reviewer 3}  
    \subsection{Main Concerns}
		\begin{enumerate} 
			\item I am not at all sure that isotopic scaling of the multi-resolution is appropriate for brain connections, but I cannot say at this point that it is not.

			\emph{It is generally accepted that the intracortical connectivity structure is isotropic. However, as the reviewer pointed out, this is not the case for cortico-cortical connectivity. Nevertheless, the presented framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions (that can form a net anisotropic kernel). The connectivity kernel can take any arbitrary shapes but this may require more basis functions for an accurate reconstruction.}

			\item This model cries out for validation in real neuronal systems, but alas, at this point we only see convergence to synthetic data.

			 \emph{We agree with this statement 100 percent. We hope that this paper will motivate new experiments, where neural field models can be experimentally validated. It has now been nearly 40 years since neural field models where introduced in computational studies. For the first time we are at the exciting point in time where computing power and estimation algorithms are sufficiently advanced to link models to actual data. This is why we believe the framework that is introduced in this paper has the potential to make an impact. For further remarks regarding this point please refer to Response 2 to the handling editor.}

### 2012-08-22 Multiresolution_framework/ltx/response_to_reviews.tex

\emph{The use of the RTS smoother in the E-step of the EM algorithm for linear state-space models is an effective, standard and widely used approach. In fact, the aim of the IDE representation in a state-space form is to adopt the EM and consequently the RTS smoother in the E-step. The RTS algorithm is presented for completeness to interested readers.}

		\emph{For linear state-space models, under Gaussian conditions, the lower bound on the log-likelihood function is always quadratic in parameters, and can be solved analytically when the parameters are separable giving results equivalent to the least square estimates. In fact this is one of the advantages of this representation, such a closed form solution for the system's parameters results in significant computational savings.}

        \item Appendix A is the analogue of Appendix D of Freestone et al. and Appendix C includes results known from the literature (as the authors point out).

\emph{In order to formulate the state-space representation of the multi-resolution neural field model, the inner product of the B-spline functions are required. We believe that it is necessary to provide the interested reader with the required tools to implement the algorithm. Appendix D of Freestone et al. (2011) and Appendix A herein are only similar in the sense that the inner product of two basis functions are calculated. In Appendix D of Freestone et al. (2011), two  n-dimensional Gaussians are considered, while here the inner product of two B-spline functions is calculated. We have no objections to removing Appendix C (now Appendix D). However, as stated above, we think that it would be informative for the broad audience of NeuroImage to include the model-specific derivations and some of the well-known formulations of the EM algorithm, since it draws on a wide range of literature.}

@@

\emph{Within our framework, the alternate levels are defined by the spatial scales of the wavelet decomposition. The significant result is the formulation of a model when various spatial scales of a neural field can be simultaneously represented. The ability to image the multi-dimensional neural field will lead to a better understanding of the important spatial scales when neural information processing occurs.}

		\item Similar claims to the ones mentioned here regarding implications for ``patient-specific'' modeling have also been made in Freestone et al. (2011), so it is not clear how these relate to the current multi-resolution approach and also such remarks might look too ambitious given that only synthetic data with fixed parameters are considered.

	\emph{Please refer to Response 2 to the handling editor. In summary, the framework can be used to create patient-specific models when it is applied with patient-specific data. Demonstrating convergence with synthetic data, where actual states and parameters are known, is a necessary first step. Furthermore, there is a significant body of work where various phenomena have been replicated using the neural field equations. This motivates the creation of tools to verify the models and experimentally test hypotheses that computational studies have generated.}

	\item The comments about model complexity and parsimony on p.13 (now p.12) are somehow vague as these quantities are undefined.

@@

		\emph{To clarify this comment, a short paragraph has been included in Section 6 to better explain the modeling choices facing every user. That is, choosing between a more parsimonious or more complex model structure.  For the audience with less experience in system identification, a reference with further information regarding this topic has also been included.}

		\item The comments on large scale networks and delays made on p.20 (now p.21) look irrelevant to the present framework.

### 2012-08-22 Multiresolution_framework/ltx/manuscript.tex
minor edits
general edits
cleaning up the manuscript
minor and last correction

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
 \usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
% \usepackage{hyperref}
 % \hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
 \newcommand{\dean}[1]{\textcolor{green}{#1}}
 \newcommand{\parham}[1]{\textcolor{blue}{#1}}
 \newcommand{\ken}[1]{\textcolor{red}{#1}}

@@

% To demonstrate the MRAIDE estimation framework we shall consider the parameters to be static and we shall also assume that the relationship between the mean membrane potential and the mean firing rate is predominantly linear. Note that there is a difference between forward and inverse modeling when making these assumptions. For example, by using a linear activation function in the estimator we are not assuming that there is necessarily a linear relationship between the membrane voltage and firing rate, but that the statistics of the neural masses fall within a linear operating region over the estimation time period. The major benefit of modeling the relationship between the mean membrane potential and firing rate is the ability to apply the EM algorithm. The loss in model accuracy may be offset estimation accuracy.

% The simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

Nevertheless, when describing a system, a choice needs to be made regarding the complexity of the mathematical description to be adopted. Such a choice is usually based on the principle of parsimony \citep{burnham2002}, that is; choosing the simplest model capable of capturing the system behavior of interest. Such a choice in favour of a simpler model has the added advantages of a smaller parameter space and therefore more accurate estimates and computationally efficient methods.

@@

In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. The basis structure (i.e., non-redundant versus a (tight) frame representation) is a requirement for the multi-resolution framework. An alternative to using the B-spline wavelets are Gaussian basis functions, which were adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity (ability to represent a zero spatial frequency field) of these basis functions enable the reconstruction of the neural field when it saturates. This may be beneficial to model neurological disorders such as epilepsy. Attaining fine-scale approximation of the neural field using Gaussian basis functions results in an ill-posed problem as the number of states is increased significantly. However, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \citet{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function. Therefore, these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
 \usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
% \usepackage{hyperref}
 % \hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}

@@

Nevertheless, when describing a system, a choice needs to be made regarding the complexity of the mathematical description to be adopted. Such a choice is usually based on the principle of parsimony \citep{burnham2002}, that is; choosing the simplest model capable of capturing the system behavior of interest. Such a choice in favor of a simpler model has the added advantages of a smaller parameter space and therefore more accurate estimates and computationally efficient methods.

@@

In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. An alternative to using the B-spline wavelets are Gaussian basis functions, which were adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity (ability to represent a zero spatial frequency field) of these basis functions enable the reconstruction of the neural field when it saturates. This may be beneficial to model neurological disorders such as epilepsy. Attaining fine-scale approximation of the neural field using Gaussian basis functions results in an ill-posed problem as the number of states is increased significantly. However, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \citet{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function. Therefore, these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.
